---
title: "Creating custom annotations"
output: html_document
---

```{r opts}
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618,
  fig.retina = 2, dpi = 150
)
```

```{r packages}
library(tidyverse)
library(here)
library(WDI) # Get data from the World Bank
library(ggrepel) # For non-overlapping labels
library(ggtext) # For fancier text handling
```

## Annotations

For this example, we're again going to use cross-national data from the [World Bank's Open Data portal](https://data.worldbank.org/). We'll download the data with the [**WDI** package](https://cran.r-project.org/web/packages/WDI/index.html).

### Load data

```{r get-wdi-fake, include = FALSE, eval = FALSE}
indicators <- c(
  "SP.POP.TOTL", # Population
  "EN.ATM.CO2E.PC", # CO2 emissions
  "NY.GDP.PCAP.KD"
) # GDP per capita

wdi_co2_raw <- WDI(
  country = "all", indicators, extra = TRUE,
  start = 1995, end = 2015
)
```

```{r save-wdi, eval = FALSE, include = FALSE}
write_csv(wdi_co2_raw, here::here("data", "wdi_co2.csv"))
```

```{r load-data-wdi-real}
wdi_co2_raw <- read_csv(here("data", "wdi_co2.csv"))
```

Then we clean the data by removing non-country countries and renaming some of the columns.

```{r clean-data-wdi}
wdi_clean <- wdi_co2_raw %>%
  filter(region != "Aggregates") %>%
  select(iso2c, iso3c, country, year,
    population = SP.POP.TOTL,
    co2_emissions = EN.ATM.CO2E.PC,
    gdp_per_cap = NY.GDP.PCAP.KD,
    region, income
  )
```

### Clean and reshape data

Next we'll do some substantial filtering and reshaping so that we can end up with the rankings of CO~2~ emissions in 1995 and 2014. I annotate as much as possible below so you can see what's happening in each step.

```{r rearrange-data}
co2_rankings <- wdi_clean %>%
  # Get rid of smaller countries
  filter(population > 200000) %>%
  # Only look at two years
  filter(year %in% c(1995, 2014)) %>%
  # Get rid of all the rows that have missing values in co2_emissions
  drop_na(co2_emissions) %>%
  # Look at each year individually and rank countries based on their emissions that year
  group_by(year) %>%
  mutate(ranking = rank(co2_emissions)) %>%
  ungroup() %>%
  # Only select a handful of columns, mostly just the newly created "ranking"
  # column and some country identifiers
  select(iso3c, country, year, region, income, ranking) %>%
  # Right now the data is tidy and long, but we want to widen it and create
  # separate columns for emissions in 1995 and in 2014. pivot_wider() will make
  # new columns based on the existing "year" column (that's what `names_from`
  # does), and it will add "rank_" as the prefix, so that the new columns will
  # be "rank_1995" and "rank_2014". The values that go in those new columns will
  # come from the existing "ranking" column
  pivot_wider(names_from = year, names_prefix = "rank_", values_from = ranking) %>%
  # Find the difference in ranking between 2014 and 1995
  mutate(rank_diff = rank_2014 - rank_1995) %>%
  # Remove all rows where there's a missing value in the rank_diff column
  drop_na(rank_diff) %>%
  # Make an indicator variable that is true of the absolute value of the
  # difference in rankings is greater than 25. 25 is arbitrary here—that just
  # felt like a big change in rankings
  mutate(big_change = ifelse(abs(rank_diff) >= 25, TRUE, FALSE)) %>%
  # Make another indicator variable that indicates if the rank improved by a
  # lot, worsened by a lot, or didn't change much. We use the case_when()
  # function, which is like a fancy version of ifelse() that takes multiple
  # conditions. This is how it generally works:
  #
  # case_when(
  #  some_test ~ value_if_true,
  #  some_other_test ~ value_if_true,
  #  TRUE ~ value_otherwise
  # )
  mutate(better_big_change = case_when(
    rank_diff <= -25 ~ "Rank improved",
    rank_diff >= 25 ~ "Rank worsened",
    TRUE ~ "Rank changed a little"
  ))
```

Here's what that reshaped data looked like before:

```{r show-head-original}
head(wdi_clean)
```

And here's what it looks like now:

```{r show-head-new}
head(co2_rankings)
```

### Plot the data and annotate

I use Roboto Condensed in this plot. You can [download it from Google Fonts](https://fonts.google.com/specimen/Roboto+Condensed?query=roboto+condensed).

```{r build-pretty-plot, fig.width=10, fig.height=6}
# These three functions make it so all geoms that use text, label, and
# label_repel will use Roboto Condensed as the font. Those layers are *not*
# influenced by whatever you include in the base_family argument in something
# like theme_bw(), so ordinarily you'd need to specify the font in each
# individual annotate(geom = "text") layer or geom_label() layer, and that's
# tedious! This removes that tediousness.
update_geom_defaults("text", list(family = "Roboto Condensed"))
update_geom_defaults("label", list(family = "Roboto Condensed"))
update_geom_defaults("label_repel", list(family = "Roboto Condensed"))

set.seed(123)

ggplot(
  data = co2_rankings,
  mapping = aes(x = rank_1995, y = rank_2014)
) +
  # Add a reference line that goes from the bottom corner to the top corner
  annotate(geom = "segment", x = 0, xend = 172, y = 0, yend = 178) +
  # Add points and color them by the type of change in rankings
  geom_point(aes(color = better_big_change)) +
  # Add repelled labels. Only use data where big_change is TRUE. Fill them by
  # the type of change (so they match the color in geom_point() above) and use
  # white text
  geom_label_repel(
    data = filter(co2_rankings, big_change == TRUE),
    aes(label = country, fill = better_big_change),
    color = "white"
  ) +
  # Add notes about what the outliers mean in the bottom left and top right
  # corners. These are italicized and light grey. The text in the bottom corner
  # is justified to the right with hjust = 1, and the text in the top corner is
  # justified to the left with hjust = 0
  annotate(
    geom = "text", x = 170, y = 6, label = "Outliers improving",
    fontface = "italic", hjust = 1, color = "grey50"
  ) +
  annotate(
    geom = "text", x = 2, y = 170, label = "Outliers worsening",
    fontface = "italic", hjust = 0, color = "grey50"
  ) +
  # Use three different colors for the points
  scale_color_manual(values = c("grey50", "#0074D9", "#FF4136")) +
  # Use two different colors for the filled labels. There are no grey labels, so
  # we don't have to specify that color
  scale_fill_manual(values = c("#0074D9", "#FF4136")) +
  # Make the x and y axes expand all the way to the edges of the plot area and
  # add breaks every 25 units from 0 to 175
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 175, 25)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 175, 25)) +
  # Add labels! There are a couple fancy things here.
  # 1. In the title we wrap the 2 of CO2 in the HTML <sub></sub> tag so that the
  #    number gets subscripted. The only way this will actually get parsed as
  #    HTML is if we tell the plot.title to use element_markdown() in the
  #    theme() function, and element_markdown() comes from the ggtext package.
  # 2. In the subtitle we bold the two words **improved** and **worsened** using
  #    Markdown asterisks. We also wrap these words with HTML span tags with
  #    inline CSS to specify the color of the text. Like the title, this will
  #    only be processed and parsed as HTML and Markdown if we tell the p
  #    lot.subtitle to use element_markdown() in the theme() function.
  labs(
    x = "Rank in 1995", y = "Rank in 2014",
    title = "Changes in CO<sub>2</sub> emission rankings between 1995 and 2014",
    subtitle = "Countries that <span style='color: #0074D9'>**improved**</span> or <span style='color: #FF4136'>**worsened**</span> more than 25 positions in the rankings highlighted",
    caption = "Source: The World Bank.\nCountries with populations of less than 200,000 excluded."
  ) +
  # Turn off the legends for color and fill, since the subtitle includes that
  guides(color = "none", fill = "none") +
  # Use theme_bw() with Roboto Condensed
  theme_bw(base_family = "Roboto Condensed") +
  # Tell the title and subtitle to be treated as Markdown/HTML, make the title
  # 1.6x the size of the base font, and make the subtitle 1.3x the size of the
  # base font. Also add a little larger margin on the right of the plot so that
  # the 175 doesn't get cut off.
  theme(
    plot.title = element_markdown(face = "bold", size = rel(1.6)),
    plot.subtitle = element_markdown(size = rel(1.3)),
    plot.margin = unit(c(0.5, 1, 0.5, 0.5), units = "lines")
  )
```
