---
title: "Creating custom annotations"
output: html_document
---

```{r opts}
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618,
  fig.retina = 2, dpi = 150
)
```

```{r packages}
library(tidyverse)
library(here)
library(WDI) # Get data from the World Bank
library(ggrepel) # For non-overlapping labels
library(ggtext) # For fancier text handling
```

## Annotations

For this example, we're going to use cross-national data from the [World Bank's Open Data portal](https://data.worldbank.org/).

### Load data

```{r load-data-wdi-real}
wdi_co2_raw <- read_csv(here("data", "wdi_co2.csv"))
```

Then we clean the data by removing non-country countries and renaming some of the columns.

```{r clean-data-wdi}
wdi_clean <- wdi_co2_raw %>%
  filter(region != "Aggregates") %>%
  select(iso2c, iso3c, country, year,
    population = SP.POP.TOTL,
    co2_emissions = EN.ATM.CO2E.PC,
    gdp_per_cap = NY.GDP.PCAP.KD,
    region, income
  )
```

### Clean and reshape data

Next we'll do some substantial filtering and reshaping so that we can end up with the rankings of CO~2~ emissions in 1995 and 2014. I annotate as much as possible below so you can see what's happening in each step.

```{r rearrange-data}
co2_rankings <- wdi_clean %>%
  # Get rid of smaller countries
  filter(population > 200000) %>%
  # Only look at two years
  filter(year %in% c(1995, 2014)) %>%
  # Get rid of all the rows that have missing values in co2_emissions
  drop_na(co2_emissions) %>%
  # Look at each year individually and rank countries based on their emissions that year
  group_by(year) %>%
  mutate(ranking = rank(co2_emissions)) %>%
  ungroup() %>%
  # Only select a handful of columns, mostly just the newly created "ranking"
  # column and some country identifiers
  select(iso3c, country, year, region, income, ranking) %>%
  # Right now the data is tidy and long, but we want to widen it and create
  # separate columns for emissions in 1995 and in 2014. pivot_wider() will make
  # new columns based on the existing "year" column (that's what `names_from`
  # does), and it will add "rank_" as the prefix, so that the new columns will
  # be "rank_1995" and "rank_2014". The values that go in those new columns will
  # come from the existing "ranking" column
  pivot_wider(names_from = year, names_prefix = "rank_", values_from = ranking) %>%
  # Find the difference in ranking between 2014 and 1995
  mutate(rank_diff = rank_2014 - rank_1995) %>%
  # Remove all rows where there's a missing value in the rank_diff column
  drop_na(rank_diff) %>%
  # Make an indicator variable that is true of the absolute value of the
  # difference in rankings is greater than 25. 25 is arbitrary here—that just
  # felt like a big change in rankings
  mutate(big_change = ifelse(abs(rank_diff) >= 25, TRUE, FALSE)) %>%
  # Make another indicator variable that indicates if the rank improved by a
  # lot, worsened by a lot, or didn't change much. We use the case_when()
  # function, which is like a fancy version of ifelse() that takes multiple
  # conditions. This is how it generally works:
  #
  # case_when(
  #  some_test ~ value_if_true,
  #  some_other_test ~ value_if_true,
  #  TRUE ~ value_otherwise
  # )
  mutate(better_big_change = case_when(
    rank_diff <= -25 ~ "Rank improved",
    rank_diff >= 25 ~ "Rank worsened",
    TRUE ~ "Rank changed a little"
  ))
```

Here's what that reshaped data looked like before:

```{r show-head-original}
head(wdi_clean)
```

And here's what it looks like now:

```{r show-head-new}
head(co2_rankings)
```

### Plot the data and annotate

```{r build-pretty-plot, fig.width=10, fig.height=6}


```
